Forecasting interrupted time series

### 2024-08-19 fits.qmd
add Github link

Time series forecasting models use historical data to estimate future values [@fildes2008forecasting] based on consistent patterns observed in the past. However, time series are sometimes disrupted by unusual events that jeopardise the regularity of the time series pattern. This can be defined as an *interrupted time series*. Time series data may be disrupted by a variety of reasons, ranging from temporary policy changes to natural disasters. Outbreaks, epidemics, and pandemics, for example, may have a considerable impact on population behavior, causing discrepancies and interruptions of data in a variety of industries such as healthcare, pharmaceuticals, transportation, retail, tourism, and traffic management. Similarly, transportation-related data may be interrupted due to equipment failures, such as malfunctions or breakdowns in vehicles, aircraft, or rail systems, resulting in missing or inaccurate data points. Furthermore, cybersecurity breaches data quality by possibly skewing readings in datasets relating to financial transactions, network traffic, and user activity, complicating time series analysis. Supply chain disruptions induced by factors such as strikes or shortages cause irregularities in sales, inventory levels, production, and delivery, often resulting in data gaps. \textcolor{teal}{For example, the war in Ukraine caused disruptions in sunflower oil and wheat supply chains, which were most likely driven by social media and people's recent memories of the COVID-19 toilet paper shortage. A power outage in Eastern Germany halted production at a Tesla facility and cut off electricity to a large retailer distribution centre, disrupting supplies to around 500 stores in and around Berlin for several weeks. Recent strikes in various forms of transport, including air and rail in Germany, France, and the United Kingdom, have had an impact on passenger flow and retail sales at airports and train stations. These issues highlight the significance of robust forecasting systems in addressing such disruptions effectively.}

@@

In this section, we describe several possible strategies to handle interruptions when forecasting time series data. \textcolor{teal}{The handling of time series interruptions is a complex and multifaceted issue, and the choice of the most appropriate strategy depends on the specific characteristics of the data and the nature of the interruption. Therefore, in this paper, we propose multiple strategies rather than a single approach to effectively address different situations and cases.}

@@

\textcolor{teal}{The disadvantage of smoothing and other highly adaptive techniques is that they need time to adapt after a disturbance and can eventually end up chasing noise. The latter is an intriguing issue since individuals often prefer adaptability while failing to recognise that it does not necessarily improve forecasting.}

@@

A special case of the intervention models proposed by @box1975intervention is a dynamic regression (or regression with ARIMA errors) model. This model includes covariates describing the interruption explicitly, with an ARIMA process describing the remaaining time series dynamics. For example, if the intervention involves a simple level shift with a reverse level shift at the end of the intervention, we can use a dummy variable to indicate the interruption period and allow the model to adjust to the interruption. More complicated interventions can be handled by using more covariates.

This has the advantage of retaining the memory of the past, and so the seasonal patterns and other dynamics will be retained. This allows the change period to be effectively modelled, provided the intervention variables are chosen well. \textcolor{teal}{Intervention models are useful because they enable the forecasting of similar future disruptions, but distinguishing between a disruption and a general driving factor, such as a promotion or a seasonal storm, may become subjective.} However, the model will assume that the post-interruption period will be similar to the pre-interruption period, and so the prediction intervals may be too narrow, especially if there is a lasting effect beyond the end of the interruption.

@@

\textcolor{teal}{When forecasting interrupted time series data, exogenous variables associated with specific events should be identified and incorporated where possible, regardless of which approach is employed, since they may improve forecasting model performance.}

@@

\textcolor{teal}{Moreover, in determining which strategy to apply in each case, it is important to think through what is being assumed, and what data are being used. This is a qualitative and subjective process, and cannot simply be tackled with a quantitative comparison of forecast accuracy. The strategies we discuss have different ways of handling the disrupted period, and many of them do not seek to provide accurate forecasts during the disruption. For example, the "set to missing" strategy ignores the disrupted period, while the "estimate what might have been" strategy replaces the observations during the disruption with interpolations. In both cases, it does not make sense to evaluate the performance of the method during the disruption. Instead, the analyst needs to decide what approach makes the most sense for the application at hand.}

\textcolor{teal}{This study proposes multiple viable strategies for forecasting time series data interrupted by different type of events. Future research could focus on exploring new strategies for handling such disrupted time series, or enhancing probabilistic forecasting methods that are designed specifically for such data. Large-scale empirical research on further exploring which strategy may work best in different scenarios also provide useful insights.} Comparative studies that assess the accuracy, robustness, and computing efficiency of different techniques across several datasets and disruption situations can help discover the most effective approaches for a range of practical applications.

@@

To enhance reproducibility and facilitate the adoption of proposed strategies, we provide the data and R code, as well as the entire paper written in R using [Quarto](https://quarto.org/) and the targets package for R [@targetsr2021]. All materials required to reproduce this paper will be accessible via a public GitHub repository once the paper is accepted for publication.

@@

We thank Dr Stephan Kolassa and several referees for their thoughts and feedback on the first version of our paper. These comments have helped us to refine and improve our work.

>>>>

Time series forecasting models use historical data to estimate future values [@fildes2008forecasting] based on consistent patterns observed in the past. However, time series are sometimes disrupted by unusual events that jeopardise the regularity of the time series pattern. This can be defined as an *interrupted time series*. Time series data may be disrupted by a variety of reasons, ranging from temporary policy changes to natural disasters. Outbreaks, epidemics, and pandemics, for example, may have a considerable impact on population behavior, causing discrepancies and interruptions of data in a variety of industries such as healthcare, pharmaceuticals, transportation, retail, tourism, and traffic management. Similarly, transportation-related data may be interrupted due to equipment failures, such as malfunctions or breakdowns in vehicles, aircraft, or rail systems, resulting in missing or inaccurate data points. Furthermore, cybersecurity breaches data quality by possibly skewing readings in datasets relating to financial transactions, network traffic, and user activity, complicating time series analysis. Supply chain disruptions induced by factors such as strikes or shortages cause irregularities in sales, inventory levels, production, and delivery, often resulting in data gaps. For example, the war in Ukraine caused disruptions in sunflower oil and wheat supply chains, which were most likely driven by social media and people's recent memories of the COVID-19 toilet paper shortage. A power outage in Eastern Germany halted production at a Tesla facility and cut off electricity to a large retailer distribution centre, disrupting supplies to around 500 stores in and around Berlin for several weeks. Recent strikes in various forms of transport, including air and rail in Germany, France, and the United Kingdom, have had an impact on passenger flow and retail sales at airports and train stations. These issues highlight the significance of robust forecasting systems in addressing such disruptions effectively.

@@

In this section, we describe several possible strategies to handle interruptions when forecasting time series data. The handling of time series interruptions is a complex and multifaceted issue, and the choice of the most appropriate strategy depends on the specific characteristics of the data and the nature of the interruption. Therefore, in this paper, we propose multiple strategies rather than a single approach to effectively address different situations and cases.

@@

The disadvantage of smoothing and other highly adaptive techniques is that they need time to adapt after a disturbance and can eventually end up chasing noise. The latter is an intriguing issue since individuals often prefer adaptability while failing to recognise that it does not necessarily improve forecasting.

@@

A special case of the intervention models proposed by @box1975intervention is a dynamic regression (or regression with ARIMA errors) model. This model includes covariates describing the interruption explicitly, with an ARIMA process describing the remaining time series dynamics. For example, if the intervention involves a simple level shift with a reverse level shift at the end of the intervention, we can use a dummy variable to indicate the interruption period and allow the model to adjust to the interruption. More complicated interventions can be handled by using more covariates.

This has the advantage of retaining the memory of the past, and so the seasonal patterns and other dynamics will be retained. This allows the change period to be effectively modelled, provided the intervention variables are chosen well. Intervention models are useful because they enable the forecasting of similar future disruptions, but distinguishing between a disruption and a general driving factor, such as a promotion or a seasonal storm, may become subjective. However, the model will assume that the post-interruption period will be similar to the pre-interruption period, and so the prediction intervals may be too narrow, especially if there is a lasting effect beyond the end of the interruption.

@@

When forecasting interrupted time series data, exogenous variables associated with specific events should be identified and incorporated where possible, regardless of which approach is employed, since they may improve forecasting model performance.

@@

Moreover, in determining which strategy to apply in each case, it is important to think through what is being assumed, and what data are being used. This is a qualitative and subjective process, and cannot simply be tackled with a quantitative comparison of forecast accuracy. The strategies we discuss have different ways of handling the disrupted period, and many of them do not seek to provide accurate forecasts during the disruption. For example, the "set to missing" strategy ignores the disrupted period, while the "estimate what might have been" strategy replaces the observations during the disruption with interpolations. In both cases, it does not make sense to evaluate the performance of the method during the disruption. Instead, the analyst needs to decide what approach makes the most sense for the application at hand.

This study proposes multiple viable strategies for forecasting time series data interrupted by different type of events. Future research could focus on exploring new strategies for handling such disrupted time series, or enhancing probabilistic forecasting methods that are designed specifically for such data. Large-scale empirical research on further exploring which strategy may work best in different scenarios also provide useful insights. Comparative studies that assess the accuracy, robustness, and computing efficiency of different techniques across several datasets and disruption situations can help discover the most effective approaches for a range of practical applications.

@@

To enhance reproducibility and facilitate the adoption of proposed strategies, we provide the data and R code, as well as the entire paper written in R using [Quarto](https://quarto.org/) and the targets package for R [@targetsr2021]. All materials required to reproduce this paper is accessible via the [GitHub repository](https://github.com/bahmanrostamitabar/forecasting_interrupted_time_series) .

@@

We thank Dr. Stephan Kolassa and several referees for their thoughts and feedback on the first version of our paper. These comments have helped us to refine and improve our work.

### 2024-08-19 fits.tex
add Github link

\title{Forecasting interrupted time series}
\date{17 May 2024}

@@

\wp{no/yr}
\jel{C10,C14,C22}

@@

\nocover
  \blind
  \lfoot{\sf17 May 2024}

@@

Time series forecasting models use historical data to estimate future
values \autocite{fildes2008forecasting} based on consistent patterns
observed in the past. However, time series are sometimes disrupted by
unusual events that jeopardise the regularity of the time series
pattern. This can be defined as an \emph{interrupted time series}. Time
series data may be disrupted by a variety of reasons, ranging from
temporary policy changes to natural disasters. Outbreaks, epidemics, and
pandemics, for example, may have a considerable impact on population
behavior, causing discrepancies and interruptions of data in a variety
of industries such as healthcare, pharmaceuticals, transportation,
retail, tourism, and traffic management. Similarly,
transportation-related data may be interrupted due to equipment
failures, such as malfunctions or breakdowns in vehicles, aircraft, or
rail systems, resulting in missing or inaccurate data points.
Furthermore, cybersecurity breaches data quality by possibly skewing
readings in datasets relating to financial transactions, network
traffic, and user activity, complicating time series analysis. Supply
chain disruptions induced by factors such as strikes or shortages cause
irregularities in sales, inventory levels, production, and delivery,
often resulting in data gaps.
\textcolor{teal}{For example, the war in Ukraine caused disruptions in sunflower oil and wheat supply chains, which were most likely driven by social media and people's recent memories of the COVID-19 toilet paper shortage. A power outage in Eastern Germany halted production at a Tesla facility and cut off electricity to a large retailer distribution centre, disrupting supplies to around 500 stores in and around Berlin for several weeks. Recent strikes in various forms of transport, including air and rail in Germany, France, and the United Kingdom, have had an impact on passenger flow and retail sales at airports and train stations. These issues highlight the significance of robust forecasting systems in addressing such disruptions effectively.}

@@

A special case of the intervention models proposed by
\textcite{box1975intervention} is a dynamic regression (or regression
with ARIMA errors) model. This model includes covariates describing the
interruption explicitly, with an ARIMA process describing the remaaining
time series dynamics. For example, if the intervention involves a simple
level shift with a reverse level shift at the end of the intervention,
we can use a dummy variable to indicate the interruption period and
allow the model to adjust to the interruption. More complicated
interventions can be handled by using more covariates.

This has the advantage of retaining the memory of the past, and so the
seasonal patterns and other dynamics will be retained. This allows the
change period to be effectively modelled, provided the intervention
variables are chosen well.
\textcolor{teal}{Intervention models are useful because they enable the forecasting of similar future disruptions, but distinguishing between a disruption and a general driving factor, such as a promotion or a seasonal storm, may become subjective.}
However, the model will assume that the post-interruption period will be
similar to the pre-interruption period, and so the prediction intervals
may be too narrow, especially if there is a lasting effect beyond the
end of the interruption.

@@

\textcolor{teal}{Moreover, in determining which strategy to apply in each case, it is important to think through what is being assumed, and what data are being used. This is a qualitative and subjective process, and cannot simply be tackled with a quantitative comparison of forecast accuracy. The strategies we discuss have different ways of handling the disrupted period, and many of them do not seek to provide accurate forecasts during the disruption. For example, the "set to missing" strategy ignores the disrupted period, while the "estimate what might have been" strategy replaces the observations during the disruption with interpolations. In both cases, it does not make sense to evaluate the performance of the method during the disruption. Instead, the analyst needs to decide what approach makes the most sense for the application at hand.}

\textcolor{teal}{This study proposes multiple viable strategies for forecasting time series data interrupted by different type of events. Future research could focus on exploring new strategies for handling such disrupted time series, or enhancing probabilistic forecasting methods that are designed specifically for such data. Large-scale empirical research on further exploring which strategy may work best in different scenarios also provide useful insights.}
Comparative studies that assess the accuracy, robustness, and computing
efficiency of different techniques across several datasets and
disruption situations can help discover the most effective approaches
for a range of practical applications.

@@

To enhance reproducibility and facilitate the adoption of proposed
strategies, we provide the data and R code, as well as the entire paper
written in R using \href{https://quarto.org/}{Quarto} and the targets
package for R \autocite{targetsr2021}. All materials required to
reproduce this paper will be accessible via a public GitHub repository
once the paper is accepted for publication.

@@

We thank Dr Stephan Kolassa and several referees for their thoughts and
feedback on the first version of our paper. These comments have helped
us to refine and improve our work.

>>>>

\title{Forecasting interrupted time series}
\date{19 August 2024}

@@

\wp{no/yr}
\nojel

@@

\nocover
  \author{Rob J Hyndman, Bahman Rostami-Tabar}
  \addresses{%
    %
      \textbf{Rob J Hyndman}\\%
      %
        Department of Econometrics \& Business Statistics\\%
        %
        %
        %
      %
        %
        Monash University\\%
        %
        %
      %
        %
        %
        Clayton VIC \\%
        %
      %
        %
        %
        %
        Australia\\%
      %
        %
        %
        %
        %
      %
      {Email: Rob.Hyndman@monash.edu}\\%
      \\[0.5cm]%
   %
      \textbf{Bahman Rostami-Tabar}\\%
      %
        %
        Cardiff Business School\\%
        %
        %
      %
        %
        %
        %
        %
      %
        %
        %
        %
        United Kingdom\\%
      %
      {Email: rostami-tabarb@cardiff.ac.uk}\\%
      \textit{Corresponding author}\\[0.5cm]%
   %
   }%
   \lfoot{\sf Hyndman, Rostami-Tabar: 19 August 2024}

@@

Time series forecasting models use historical data to estimate future
values \autocite{fildes2008forecasting} based on consistent patterns
observed in the past. However, time series are sometimes disrupted by
unusual events that jeopardise the regularity of the time series
pattern. This can be defined as an \emph{interrupted time series}. Time
series data may be disrupted by a variety of reasons, ranging from
temporary policy changes to natural disasters. Outbreaks, epidemics, and
pandemics, for example, may have a considerable impact on population
behavior, causing discrepancies and interruptions of data in a variety
of industries such as healthcare, pharmaceuticals, transportation,
retail, tourism, and traffic management. Similarly,
transportation-related data may be interrupted due to equipment
failures, such as malfunctions or breakdowns in vehicles, aircraft, or
rail systems, resulting in missing or inaccurate data points.
Furthermore, cybersecurity breaches data quality by possibly skewing
readings in datasets relating to financial transactions, network
traffic, and user activity, complicating time series analysis. Supply
chain disruptions induced by factors such as strikes or shortages cause
irregularities in sales, inventory levels, production, and delivery,
often resulting in data gaps. For example, the war in Ukraine caused
disruptions in sunflower oil and wheat supply chains, which were most
likely driven by social media and people's recent memories of the
COVID-19 toilet paper shortage. A power outage in Eastern Germany halted
production at a Tesla facility and cut off electricity to a large
retailer distribution centre, disrupting supplies to around 500 stores
in and around Berlin for several weeks. Recent strikes in various forms
of transport, including air and rail in Germany, France, and the United
Kingdom, have had an impact on passenger flow and retail sales at
airports and train stations. These issues highlight the significance of
robust forecasting systems in addressing such disruptions effectively.

@@

A special case of the intervention models proposed by
\textcite{box1975intervention} is a dynamic regression (or regression
with ARIMA errors) model. This model includes covariates describing the
interruption explicitly, with an ARIMA process describing the remaining
time series dynamics. For example, if the intervention involves a simple
level shift with a reverse level shift at the end of the intervention,
we can use a dummy variable to indicate the interruption period and
allow the model to adjust to the interruption. More complicated
interventions can be handled by using more covariates.

This has the advantage of retaining the memory of the past, and so the
seasonal patterns and other dynamics will be retained. This allows the
change period to be effectively modelled, provided the intervention
variables are chosen well. Intervention models are useful because they
enable the forecasting of similar future disruptions, but distinguishing
between a disruption and a general driving factor, such as a promotion
or a seasonal storm, may become subjective. However, the model will
assume that the post-interruption period will be similar to the
pre-interruption period, and so the prediction intervals may be too
narrow, especially if there is a lasting effect beyond the end of the
interruption.

@@

Moreover, in determining which strategy to apply in each case, it is
important to think through what is being assumed, and what data are
being used. This is a qualitative and subjective process, and cannot
simply be tackled with a quantitative comparison of forecast accuracy.
The strategies we discuss have different ways of handling the disrupted
period, and many of them do not seek to provide accurate forecasts
during the disruption. For example, the ``set to missing'' strategy
ignores the disrupted period, while the ``estimate what might have
been'' strategy replaces the observations during the disruption with
interpolations. In both cases, it does not make sense to evaluate the
performance of the method during the disruption. Instead, the analyst
needs to decide what approach makes the most sense for the application
at hand.

This study proposes multiple viable strategies for forecasting time
series data interrupted by different type of events. Future research
could focus on exploring new strategies for handling such disrupted time
series, or enhancing probabilistic forecasting methods that are designed
specifically for such data. Large-scale empirical research on further
exploring which strategy may work best in different scenarios also
provide useful insights. Comparative studies that assess the accuracy,
robustness, and computing efficiency of different techniques across
several datasets and disruption situations can help discover the most
effective approaches for a range of practical applications.

@@

To enhance reproducibility and facilitate the adoption of proposed
strategies, we provide the data and R code, as well as the entire paper
written in R using \href{https://quarto.org/}{Quarto} and the targets
package for R \autocite{targetsr2021}. All materials required to
reproduce this paper is accessible via the
\href{https://github.com/bahmanrostamitabar/forecasting_interrupted_time_series}{GitHub
repository} .

@@

We thank Dr.~Stephan Kolassa and several referees for their thoughts and
feedback on the first version of our paper. These comments have helped
us to refine and improve our work.
